Simulating High-Dimensional Multivariate Data

### 2021-03-01 010-background.Rmd
revised through 020

<!-- 
The next section describes the motivating data set we use to provide example applications.
-->

@@

Simulating high-dimensional, non-normal, correlated data motivates this work --- in pursuit of modeling RNA-sequencing (RNA-seq) data [@Wang2009b; @Conesa2016b] derived from breast cancer patients. The RNA-seq data-generating process involves counting how often a particular form of human messenger RNA (mRNA) is expressed in a biological sample. RNA-seq platforms typically quantify the entire transcriptome in one experimental run, resulting in high-dimensional data. For human-derived samples, this results in count data corresponding to over 20,000 genes (protein-coding genomic regions) or even over 77,000 isoforms when alternatively-spliced mRNA are counted [@Schissler2019]. Importantly, due to inherent biological processes, gene expression data exhibit correlation --- co-expression --- across genes [@BE07; @Schissler2018].

@@

*Correspondence among Pearson, Spearman, and Kendall correlations*.
There is no closed form, general correspondence among the rank-based measures and the Pearson correlation coefficient, as the marginal distributions $F_i$ are intrinsic in their calculation. For *bivariate normal vectors*, however, the correspondence is well-known:

@@

*Discrete marginal considerations*.
Spearman's correlation for discrete marginal suffers issues whenever there is a nonzero probability of ties. For example, the Spearman's correlation of a discrete-valued random variable $X$ with itself could be less than 1 [@MB13]. One remedy for this issue is to rescale Equation \@ref(eq:spearman). For two random variables $X,Y$ with probability mass functions (PMFs) or probability densities functions (PDFs) $p(x)$ and $q(y)$, respectively, define the rescaled Spearman's correlation as

@@

*Marginal-dependent bivariate correlation bounds*. 
Given two marginal distributions, $\rho_P$ is not free to vary over the entire range of possible correlations $[-1,1]$. The so-called *Frechet-Hoeffding bounds* are well-studied [@Nelsen2007; @BF17]. These constraints cannot be overcome through algorithm design. In general, the bounds are given by

>>>>

@@

Simulating high-dimensional, non-normal, correlated data motivates this work --- in pursuit of modeling RNA-sequencing (RNA-seq) data [@Wang2009b; @Conesa2016b] derived from breast cancer patients. The RNA-seq data-generating process involves counting how often a particular form of human messenger RNA (mRNA) is expressed in a biological sample. RNA-seq platforms typically quantify the entire transcriptome in one experimental run, resulting in high-dimensional data. For human-derived samples, this results in count data corresponding to over 20,000 genes (protein-coding genomic regions) or even over 77,000 isoforms when alternatively-spliced mRNA are counted [@Schissler2019]. Importantly, due to inherent biological processes, gene expression data exhibit correlation (co-expression) across genes [@BE07; @Schissler2018].

@@

*Correspondence among Pearson, Spearman, and Kendall correlations*

There is no closed form, general correspondence among the rank-based measures and the Pearson correlation coefficient, as the marginal distributions $F_i$ are intrinsic in their calculation. For *bivariate normal vectors*, however, the correspondence is well-known:

@@

*Discrete marginal considerations*

Spearman's correlation for discrete marginal suffers issues whenever there is a nonzero probability of ties. For example, the Spearman's correlation of a discrete-valued random variable $X$ with itself could be less than 1 [@MB13]. One remedy for this issue is to rescale Equation \@ref(eq:spearman). For two random variables $X,Y$ with probability mass functions (PMFs) or probability densities functions (PDFs) $p(x)$ and $q(y)$, respectively, define the rescaled Spearman's correlation as

@@

*Marginal-dependent bivariate correlation bounds*

Given two marginal distributions, $\rho_P$ is not free to vary over the entire range of possible correlations $[-1,1]$. The so-called *Frechet-Hoeffding bounds* are well-studied [@Nelsen2007; @BF17]. These constraints cannot be overcome through algorithm design. In general, the bounds are given by
